---
title: Spatial Objects
---

## Lesson Goals

* Understand R representation of vector and raster spatial data
* Familiarity with R packages for reading, writing, and dealing with vector and raster data
* Understand spatial file types supported by R spatial packages
* Learn to pull spatial data from online sources into R

### Spatial Data Types
#### Simple Representation of Spatial Data and Supported Data Types

We can represent spatial data as discrete locations (points, lines or polygons) or as continuous phenomena without natural boundaries. We typiclaly represent the former type of dta as *vector* data, with an associated geometry or shape, and some attributes with information about the locations.  Examples are things like state boundaries with state name and population, rivers with their flow volume and names, polygons of watersheds with their names and associated landscape information.  We represent the latter type of data (continuous spatial fields) with *rasters* - these would be things such as elevation, precipitation, and atmospheric deposition for instance.


Basic data structures in R can represent spatial data - all we need is some vectors with location and attribute information - below we generate cites with population and make a map with a legend:

```{r cities}
cities <- c('Ashland','Corvallis','Bend','Portland','Newport')
longitude <- c(-122.699, -123.275, -121.313, -122.670, -124.054)
latitude <- c(42.189, 44.57, 44.061, 45.523, 44.652)
population <- c(20062,50297,61362,537557,9603)
locs <- cbind(longitude, latitude) 
plot(locs, cex=sqrt(population*.0002), pch=20, col='red', 
  main='Population', xlim = c(-124,-120.5), ylim = c(42, 46))
text(locs, cities, pos=4)

# Add a legend
breaks <- c(20000, 50000, 60000, 100000)
options(scipen=3)
legend("topright", legend=breaks, pch=20, pt.cex=1+breaks/20000, 
  col='red', bg='gray')
```

We can add a random polygon and map this as well:
```{r}
lon <- c(-123.5, -123.5, -122.5, -122.670, -123)
lat <- c(43, 45.5, 44, 43, 43)
x <- cbind(lon, lat)
polygon(x, border='blue')
lines(x, lwd=3, col='red')
points(x, cex=2, pch=20)
```

### Vector Data with sp

### Vector Data with sf

### Raster Data

### Getting Data 
#### Downloading Open Data and Geographic Data Packages

### Geographic Data I/O